---
title: "ESS 3500 R Handbook, Lesson 4"
author: "Emily Schultz"
date: "2023-02-15"
output: 
  html_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)
```

## Lesson 4: DISTRIBUTIONS AND DESCRIPTIVE STATISTICS

In this lesson, we will cover how to graph distributions in R and how to calculate descriptive statistics to describe those distributions. The good news is, you have already learned some of the descriptive statistics functions, so they will be a review. You will also get to make your first graphs using the ggplot2 package!

We will work with the **BKSnow.csv** file, which has monthly snowfall data for Breckenridge, dating back to the late 1940s. Download the data file and upload that file into R now (be sure to set your working directory first, and save your code in a script file):

```{r load_data}
snow <- read.csv("BKSnow.csv")
```

### 4.1 Graphing distributions
In this section, we will make our first foray into graphing! We will use the **ggplot2** package (part of the **tidyverse** set of packages) for making graphs because it is easier to make nice looking graphs in ggplot than in base R.

Load the **tidyverse** package for this session:

```{r load_ggplot}
library(tidyverse)
```

In this section, you will learn to make histograms. Making this type of graph is often one of the first steps in exploratory data analysis, which gives us a feel for what our data look like before we start formal analysis.

Histograms are useful for visualizing how our data are distributed. Are they close to a normal (bell-shaped) distribution? Are they skewed? Biomodal? Are there outliers? All of these questions can be answered with a histogram.

We will start by making a histogram of our "JULJAN" variable, which is the total snowfall from July to January. When you use the **ggplot2** package to make graphs, you will start with one line of code that tells ggplot what data to use for your graph. Then you add additional lines of code to create and modify your graph. Here is the code to make a simple histogram, without changing any default settings:

```{r histogram}
ggplot(data = snow, aes(x = JULJAN)) +
  geom_histogram() 
```

In the first line of code above, we start with the `ggplot` function. The first argument of the function (data) is the data frame that contains the data you want to graph (it needs to be a data frame object). We want to use our `snow` dataframe. Next is the aesthetics argument (aes), where you list the variables you want to graph. Notice that "aes" is followed by parentheses. This is because you can list multiple variable for this argument, so you put the list of variable inside the parentheses. For our histogram, we are only using one variable (JULJAN), and this is our x variable. We end the line with a + sign to indicate that we are adding additional code on the next line to create the graph. In the second line of code, we are telling ggplot that we want to use the data from the first line of code to create a histogram. We don't need to add any arguments to the `geom_histogram` because we already told ggplot what data to use, and we aren't changing any default settings.

Now let's take a look at a few ways we can modify our graph to make it look nicer. First, let's change our axis labels. "JULJAN" is sensible for a column heading because it's brief and it doesn't have any spaces, but it' looks a little odd on a graph's not descriptive enough for an axis label. I also like to start my axis titles with a capital letter. We will add another line to the code with the `labs` function to change the axis labels. Don't forget to add a plus sign to the end of the geom_histogram line to indicate that you are adding more code for the graph below.

```{r axis_labels}
ggplot(data = snow, aes(x = JULJAN)) +
  geom_histogram() +
  labs(x = "Snowfall from July to January", y = "Frequency")
```

I also find the grey background to be distracting, so we will change the theme to get rid of the background. Again, we will add another line of code, so add the plus sign to the end of the labs line.

```{r theme}
ggplot(data = snow, aes(x = JULJAN)) +
  geom_histogram() +
  labs(x = "Snowfall from July to January", y = "Frequency") +
  theme_classic()
```

Now we have a histogram that both shows us the distribution of our data and looks pretty nice!

### 4.2 Descriptive statistics

#### Central tendency
In class, we discussed three measures of central tendency: mean, median, and mode. Below, we will cover how to calculate each for the "JULJAN" variable in the data frame.

You have used the `mean` function before, but this time we will calculate the mean of a column in  data frame instead of the mean of a vector. Notice the use of the optional `na.rm=T` argument to remove missing values before calculating the mean.

```{r mean}
mean_JULJAN <- mean(snow$JULJAN,na.rm=T)
mean_JULJAN
```

Calculating the median is similarly intuitive:

```{r median}
median_JULJAN <- median(snow$JULJAN,na.rm=T)
```

Unfortunately, there is not a built-in function for calculating the mode in R, but there are a few options for determining the mode. Likely the easiest is to install **DescTools** package, which does have a mode function.

```{r instsall_DescTools, eval = F}
install.packages("DescTools")
```

You will only need to run the install packages command once, and then it will be installed on your computer.

To load the package and calculate the mode, run

```{r mode}
library(DescTools)
mode_JULJAN <- Mode(snow$JULJAN)
mode_JULJAN
```

In the output, the top number you will see is the mode. If there are multiple modes, you will see a vector listing all of the modes instead of a single number. The number on the bottom is the frequency of the mode (how often that value appeared in the data set). Notice in this case, the values are both `NA`. Based on the variable type for the JULJAN variable and the definition of the mode, why do you think this is? What type of variable might be more likely to have a mode?

The mode is the measure of central tendency we will use the least in this class, and it is often enough to infer it visually from a frequency histogram.

Next, we will add lines representing the mean and median to the histogram we made above. This will allow us to visualize where these values fall relative to the distribution of our data. We can do this by adding two lines to the codes we already wrote to create our histogram:

```{r center_visual}
ggplot(data = snow, aes(x = JULJAN)) +
  geom_histogram() +
  geom_vline(aes(xintercept=mean_JULJAN),color="blue",linewidth=1) +
  geom_vline(aes(xintercept=median_JULJAN),color="orange",linewidth=1) +
  labs(x = "Snowfall from July to January", y = "Frequency") +
  theme_classic()
```

The two lines we added are the lines using the `geom_vline` function, which adds a vertical line to the graph. The first argument for this function is the aes argument, like we saw in the `ggplot` function. To add a vertical line, we need to give R the x-intercept of that line. In this case, we want to add lines representing the mean and median, so the intercepts of the lines will be the mean and median of the "JULJAN" variable that we calculated above. After the aes argument, I also add to optional arguments: color and linewidth. The color argument allows you to change the line color. You can use either the name of the color or the hex code. The linewidth function allows you to change the thickness of the line. Here, by setting the linewidth to 1, we are making the lines thicker than the default, so they are easier to see. 

Based on the mean and median, would you say the data are normally distributed or skewed (i.e., log normal)?

Let's see how this affects our interpretation of current snowfall and how it compares to what is "normal". We will pull out the July to January snowfall from this year and see how it compares to the mean and median.

First, we will extract and save this year's snowfall. This will come from the "JULJAN" variable in our snow data frame, so we can extract that column using the $ symbol. Then we will want to take only the last value from that column. We can do that using [] to index the column. We could count the number of rows in the data frame and use that number to extract the last value, but there is a shortcut. The `nrow` function will count the number of rows in a data frame for use, so we can just use that function inside the brackets to calculate the correct index.

The full code for this looks like this:

```{r snow_2024}
snow_2024 <- snow$JULJAN[nrow(snow)]
```

Now that we have the value, we can add it to our histogram, again using the `geom_vline` function, to visually compare it to the mean and median:

```{r compare}
ggplot(data = snow, aes(x = JULJAN)) +
  geom_histogram() +
  geom_vline(aes(xintercept=mean_JULJAN),color="blue",linewidth=1) +
  geom_vline(aes(xintercept=median_JULJAN),color="orange",linewidth=1) +
  geom_vline(aes(xintercept=snow_2024),color="black",linewidth=1) +
  labs(x = "Snowfall from July to January", y = "Frequency") +
  theme_classic()
```

Based on this new graph, is our current July to January snowfall lower than normal or higher than normal?

#### Variation

In this section, you will practice calculating measures of variability in a data set: standard deviation (which you have already learned) and variance (which is just as easy).

We'll continue working with the "JULJAN" and "LOG_JULJAN" variables in our data frame.

Calculate the standard deviation and variance as follows:

```{r variability}
sd_JULJAN <- sd(snow$JULJAN,na.rm=T)
sd_JULJAN
var_JULJAN <- var(snow$JULJAN,na.rm=T)
var_JULJAN
```

Notice the relationship between the two. The standard deviation is the square root of the variance, so we could also calculate one from the other.

### 4.3 Adding probability distributions
So far, we have been working with frequency distributions of our data, but we have been discussing using formal statistical distributions to represent our data. Now we will practice calculating a formal distribution (the normal distribution) and comparing it to our data.

R has a function, `dnorm`, that can be used to calculte the probability density of a value for a particular outcome. We can use this function to graph the probability density function for snowfall in Breckenridge by using the function to calculate the probability density for a wide range a values and graphing the output.

Based on the histogram we made, a good range of values to use is 1 to 200 because that covers the full range of snowfall values that we observed. When we use the `dnorm` function, we will provide this range as the first argument. We can create this sequence of numbers using the `:` symbol. Then, to calculate the probability density of all of these snowfall values, we have to provide the mean and standard deviation of our distribution as the next two arguments. We will use the mean and standard deviation we calculated for our JULJAN variable above. We will save the probability densities values that we calculate in a vector called "pdens".

```{r normal}
pdens <- dnorm(1:200,mean=mean_JULJAN,sd=sd_JULJAN)
```

In order to graph our data using ggplot, we have to put the data in a data frame. We will create a data frame with two variables: one with the range of snowfall values and the second with the probability densities of those snowfall values.

```{r normal_data}
pdens_data <- data.frame(snow=1:200,pdens=pdens)
```

Finally, we can graph this distribution on top of our histogram, so we can compare the distribution of our data with the probability distribution.

We will do this by adding another layer to our graph, with the `geom_density` function. In this function, we are using a different data set that the one we used for the rest of our graph ("pdens_data" instead of "snow). Because of this, inside this function, we have to tell R the new data set we want to use (first argument) and the values we want to graph (x and y inside the aes argument). In this case we want to graph the probability density for each snowfall value. Then we include another argument (stat = "identity"), which tells R that we want to graph the probability density based on the value we provide in our data frame, rather that calculating the density based on the frequency of values in the data frame. Finally, the last argument again changes the thickness of the line.

Notice I have also modified the `geom_histogram` function. I added an argument (`aes(y=..density..)`). This change rescales to values in the histogram so they represent a probability density, rather than a frequency. This will put the values on the same scale as the probability density function we calculated, so we can effectively compare the distribution of our data to the probability distribution.

Here's the full code for the graph:

```{r normal_graph}
ggplot(snow,aes(x=JULJAN)) +
  geom_histogram(aes(y=..density..))+
  geom_vline(aes(xintercept=mean_JULJAN),color="blue",linewidth=1) +
  geom_vline(aes(xintercept=median_JULJAN),color="orange",linewidth=1) +
  geom_density(data=pdens_data,aes(x=snow,y=pdens),stat="identity",linewidth=1)+
  theme_classic()
```

### Team challenge
Throughout this exercise, we worked the the JULJAN variable, which is the total snowfall from July to January. One thing we noticed is that there is a little bit of skew in this variable. In class, I mentioned that we can sometimes force a skewed distribution to be normal by taking the log of the variable. We will try that out here and see how it changes our exploratory analysis of historical snowfall and how this year's snowfall compares, following the steps below.

1. Use the `mutate` function to calculate the log of the JULJAN variable and add it to the "snow" data frame.

2. Calculate the mean and median of the log JULJAN variable.

3. Extract the value of the log JULJAN variable for 2023-2024.

4. Use the `dnorm` function to calculate the probability densities of a range of values for the log JULJAN variable. I would recommend a range from 2.5 to 6. Because this is a small range, you will get a better-looking curve if your sequence of values uses increments of 0.1 instead of 1. You can do this with the `seq` function as follows: `seq(2.5,6,0.1)`. Just use this in place of the `1:200` in the code you practiced above.

5. Create a data frame with the range of snowfall values and the probability densities that you calculated above.

6. Make a histogram showing the distribution of the log JULJAN data, the probability distribution, the mean and median of the log JULJAN data, and the 2023-2024 value of the log JULJAN data. 
