---
title: "ESS 3500 R Handbook, Lesson 14"
author: "Emily Schultz"
date: "2023-02-15"
output: 
  html_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)
```

## Lesson 14: Linear Regression
In this lesson, you will tackle a test that can be used when you have two continuous variables: linear regression. We will work with a plant cover data set that looks at climate variables as possible predictors of native plant cover.


### 14.1 Simple linear regression
In this first section, we will individually test the effect of mean temperature and maximum temperature on the cover of native plant species. To begin, we will build and visualize the models.

First, load the data set. Be sure your working directory is set correctly.

```{r plant_data}
plant <- read.csv("PlantSumm.csv")
```

For now, we'll build three models: the null model, an alternative model for mean temperature, and an alternative model for maximum temperature.

```{r plant_lm}
plant_null <- lm(tot_cover ~ 1, plant)
plant_mean <- lm(tot_cover ~ Mean_tempC, plant)
plant_max <- lm(tot_cover ~ Max_tempC, plant)
```

To view the output of your models, type the name of each model. Just like when we have worked with this model before, the null model will have just an intercept, and the alternative model will have a intercept and a slope term for the effect of either mean or maximum temperature on plant cover.

Now, let's create graphs to visualize our data. Because we have two continuous variables, a scatterplot is a good option. We will also include a best fit line based on our alternative model. We'll make two graphs, one for each of our independent variables. If you haven't already, load the `ggplot2` package first.

```{r plant_scatter}
library(ggplot2)
ggplot(plant, aes(x=Mean_tempC, y=tot_cover)) +
  geom_point() +
  geom_smooth(method="lm")+
  labs(x="Mean temperature (C)", y="Total plant cover") +
  theme_classic()

ggplot(plant, aes(x=Max_tempC, y=tot_cover)) +
  geom_point() +
  geom_smooth(method="lm")+
  labs(x="Max temperature (C)", y="Total plant cover") +
  theme_classic()
```

#### Classical Frequentist Approach
Let's again start by analyzing the models using a frequentist approach. We don't actually need to run any additional tests for this. We can just look at some additional output from the models we already ran.

To view the additional output, use the `summary` function.

```{r lm_out}
summary(plant_mean)
summary(plant_max)
```

When you view the output, you will see a number of things. First, you will be able to see the the formula you used to build the models. Then you will see some information on the distribution of the residuals (the leftover variation not explained by your model). Next, you will see the coefficients from your model, along with standard error of the estimates. The coefficients section will also show you t-values and p-values for each coefficient. These are one-sample t-tests comparing the value of the coefficient to zero.

The information we really want for our linear regression test is down at the very bottom. In the final section, you will see some R-squared values. These are a measure of how much variation in your dependent variable is explained by your independent variable (we will talk about this more in class). Below that, you will see the output of the linear regression test. First is the F-statistic (the same statistic that was calculated for the ANOVA). Then you will see the p-value. Based on these values, would your reject or accept the null hypothesis? What does this tell you about the effect of temperature on plant growth?

#### Likelihood-based approach
Next we will use a likelihood-based approach to test the same question. Once again, the approach is the same as what you used for the t-test and ANOVA, using the `AIC` function to compare the models. Since we have three models, we can include all three in the AIC function and then compare the AIC from both the plant_mean and plant_max models to the null model.

```{r cone_aic}
AIC(plant_null, plant_mean, plant_max)
```

Based on this output, what would you conclude about the effect of mean and maximum temperature on plant growth?


